---
title: 'Species Persistence: Data Processing and Visualization'
author: "NOAA Fisheries' Office of Science and Technology"
date: "2025-12-30"
output:
  html_document:
    toc: yes
    toc_float: true
    theme: readable
    code_folding: hide
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)

library(here)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(readr)
library(tidyverse)
library(broom)
```


## Background

This document presents various analyses for evaluating species persistence as a new analysis module on [DisMAP](https://apps-st.fisheries.noaa.gov/dismap/). 

**The Species Persistence module is in a pilot period. Feedback is encouraged and can always be given by submitting a pull request through GitHub, taking our [feedback survey](https://docs.google.com/forms/d/1yJX11wHzLBh1mQEKEKV_JcsNimdiAC75qtmqTsUw_ZE/viewform?pli=1&pli=1&edit_requested=true), or emailing us at [dismap.contact@noaa.gov](dismap.contact@noaa.gov).**


The goals for this analysis are four-fold:

* Visualize changes in observed biomass for a particular species, from year to year
* Categorize survey abundance as being: 
  + absent, 
  + rare, 
  + less common,
  + common,
  + abundant,
  + or highly abundant
* Measure overall prevalence pattern of a particular species, across all survey years
* Categorize trend as:
  + Persistently Present, 
  + Emerged,
  + Potentially Emerging,
  + Potentially Disappearing,
  + Disappeared,
  + Sporadic - Primarily Present,
  + or Sporadic - Primarily Absent

## Using this Code
*To use this code workflow*, run the following code in its entirety. To export final data sets as csv files, navigate to the two "Data Export" chunks and uncomment-out the `write.csv()` code line and rerun the section. This will export the appropriate final data sets, which also reflect the publicly displayed data on DisMAP.

```{r wrangling, message = FALSE}
### Run `Compile_Dismap_Current.R` first before this analysis ###

#read in data
data <- data.frame(base::readRDS(file = here::here("data_processing_rcode/output/data_clean/alldata_withzeros.rds"))) %>% 
  select(region, haulid, year, spp, common, wtcpue, stratum, stratumarea, lat, lon, depth)

bfish <- read_csv(here::here("data_processing_rcode/data", "BFISH_DisMAP_2024_update.csv")) %>% 
  mutate(haulid = psu,
         stratumarea = NA,
         region = case_when(region == "Hawaii" ~ "Hawai'i Islands")) %>% 
  select(region, haulid, year, spp, common, wtcpue, stratum, stratumarea, lat, lon, depth)


data <- rbind(data, bfish)
rm(bfish)
```

## Data Processing
```{r prep}
# need to edit naming for regions versus surveys
data$survey <- ifelse(data$region == "Aleutian Islands", "Aleutian Islands Bottom Trawl Survey",
                         ifelse(data$region == "Eastern Bering Sea", "Eastern Bering Sea Crab/Groundfish Bottom Trawl Survey",
                                ifelse(data$region == "Northern Bering Sea", "Northern Bering Sea Crab/Groundfish Survey - Eastern Bering Sea Shelf Survey Extension",
                                       ifelse(data$region == "Gulf of Alaska", "Gulf of Alaska Bottom Trawl Survey",
                                              ifelse(data$region == "Gulf of Mexico", "Gulf of Mexico Summer Shrimp/Groundfish Survey",
                                                     ifelse(data$region == "Northeast US Fall", "NEFSC Fall Bottom Trawl",
                                                            ifelse(data$region == "Northeast US Spring", "NEFSC Spring Bottom Trawl",
                                                                   ifelse(data$region == "Southeast US Fall", "SEAMAP Fall Coastal Trawl Survey",
                                                                          ifelse(data$region == "Southeast US Spring", "SEAMAP Spring Coastal Trawl Survey",
                                                                                 ifelse(data$region == "Southeast US Summer", "SEAMAP Summer Coastal Trawl Survey",
                                                                                        ifelse(data$region == "West Coast Annual", "West Coast Bottom Trawl Annual",
                                                                                               ifelse(data$region == "West Coast Triennial", "West Coast Bottom Trawl Triennial",
                                                                                                      ifelse(data$region == "Hawai'i Islands", "Bottomfish Fishery-Independent Survey in Hawaii (BFISH)",
                                                                                                             ifelse(data$region == "Eastern and Northern Bering Sea", NA, NA))))))))))))))

data_update <- data %>%
  filter(!is.na(survey)) %>%
  mutate(region_old = region) %>%
  select(-region)

data_update$region <- ifelse(data_update$region_old == "Aleutian Islands", "Aleutian Islands",
                             ifelse(data_update$region_old == "Eastern Bering Sea", "Eastern Bering Sea",
                                    ifelse(data_update$region_old == "Northern Bering Sea", "Northern Bering Sea",
                                           ifelse(data_update$region_old == "Gulf of Alaska", "Gulf of Alaska",
                                                  ifelse(data_update$region_old == "Gulf of Mexico", "Gulf of Mexico",
                                                         ifelse(data_update$region_old == "Northeast US Fall", "Northeast US",
                                                                ifelse(data_update$region_old == "Northeast US Spring", "Northeast US",
                                                                       ifelse(data_update$region_old == "Southeast US Fall", "Southeast US",
                                                                              ifelse(data_update$region_old == "Southeast US Spring", "Southeast US",
                                                                                     ifelse(data_update$region_old == "Southeast US Summer", "Southeast US",
                                                                                            ifelse(data_update$region_old == "West Coast Annual", "West Coast",
                                                                                                   ifelse(data_update$region_old == "Hawai'i Islands", "Hawai'i Islands",
                                                                                                          ifelse(data_update$region_old == "West Coast Triennial", "West Coast", NA)))))))))))))

```

### Biomass percentile
This section captures the data by calculating percentile/rank of biomass of a specific species, compared to all the species biomass in that survey region and survey year.
We calculate total biomass for each species in each year for each survey. Then we remove zero biomass records because these values would skew percentile calculations in the following step. These will be added back in after the percentile calculations.
```{r}
data_sum <- data_update %>%
  select(region, survey, spp, year, common, wtcpue) %>%
  group_by(region, survey, year, spp, common) %>%
  summarise(wtcpue = sum(wtcpue))

data_sum_zeros <- data_sum %>% 
  filter(wtcpue == 0)

data_sum_nozeros <- data_sum %>% 
  filter(wtcpue > 0)
```

Using the `cume_dist()` function here (as opposed to the `percent_rank()` function) to rank biomass values as percentiles because this method will not produce a percentile of zero. Percentiles are calculated for a particular species, in comparison to all the biomass values for all species in a survey/survey region, across all survey years.
```{r percentile, message = FALSE}

data_rank_nozeros <- data_sum_nozeros %>%
  group_by(region, survey) %>%
  mutate(percentile = cume_dist(wtcpue))

```

Then, we bring the zero biomass rows back in after the percentiles are calculated and `rbind` them to the data set with percentiles calculated. Species that have zero biomass in a survey region for a survey year will have a percentile value of zero. Any species that has a survey biomass > 0 for a survey year will have a ranked percentile value, in comparison to all the biomass values in that survey region across all survey years.
```{r}

data_rank <- rbind(data_rank_nozeros, data_sum_zeros) %>% 
  dplyr::rename(Region = region,
                SurveyName = survey,
                Common = common,
                Species = spp,
                Year = year,
                WTCPUE = wtcpue,
                Percentile = percentile) %>% 
  mutate(Percentile = ifelse(is.na(Percentile), 0, Percentile))
  
```

The percentile values are then binned as being "Absent" (biomass and percentile of 0), "Rare" (< 10th percentile), "Less Common" (10th - 25th percentile), "Common" (25th - 75th percentile), "Abundant" (75th - 90th percentile), or "Highly Abundant" (> 90th percentile).
```{r binning}

data_rank$Bin <- ifelse(data_rank$Percentile == 0.0, "Absent",
                        ifelse(data_rank$Percentile > 0.0 & data_rank$Percentile <= 0.10, "Rare",
                               ifelse(data_rank$Percentile > 0.10 & data_rank$Percentile <= 0.25, "Less Common",
                                      ifelse(data_rank$Percentile > 0.25 & data_rank$Percentile <= 0.75, "Common",
                                             ifelse(data_rank$Percentile > 0.75 & data_rank$Percentile <= 0.90, "Abundant",
                                                    ifelse(data_rank$Percentile > 0.90, "Highly Abundant", NA))))))

```

Then, we assign all species with survey biomass > 0 as present in that year's survey and < 0 as being absent in that year's survey.
```{r presence_absence}

data_rank$Absence_Presence <- ifelse(data_rank$WTCPUE == 0, "0",
                                     ifelse(data_rank$WTCPUE > 0, "1", "NA"))

#Absence is 0 and Presence is 1

```


### Presence in hauls
This section will calculate the number of hauls that a species is present in, per survey and per year. Measuring the presence in hauls will capture the prevalence of that species in the survey (i.e., how prominent is this species in the survey?), over time.

We calculate the number of hauls total (in each survey and survey year) and the number of hauls for which a particular species was caught. We then also calculate the proportion of total hauls each species was caught in (for each survey and survey year).

```{r build haul freq df, message = FALSE}

haulsyr_max <- data %>%
  select(survey, haulid, year) %>%
  distinct() %>%
  group_by(survey, year) %>%
  summarise(maxhauls = n())

haulsyr <- data %>% 
  filter(wtcpue > 0) %>% 
  select(survey, year, haulid, spp) %>% 
  distinct() %>% 
  group_by(survey, year, spp) %>% 
  summarise(haulspres = n())


data_haul <- left_join(haulsyr, haulsyr_max, by = c("survey", "year")) %>% 
  mutate(
    haul_perc = haulspres/maxhauls
      ) %>% 
  rename(
    SurveyName = survey,
    Year = year,
    Species = spp,
    HaulPresence = haulspres,
    HaulProportion = haul_perc
  ) %>% 
  select(-maxhauls)

```


### Binding haul count to percentiles

Using left_join to add the `haulspres` and `haul_perc` values from `data_haul` to the corresponding survey and species in `data_rank`. Any zero biomass/percentile will turn up as `NA` values in the haul presence column when it's joined. We have converted these `NA`'s to zeros because those are instances in which the species was not landed in the survey. **This builds our full data set of results.**

```{r}
data_full <- data_rank %>% 
  left_join(data_haul, by = c("SurveyName", "Year", "Species")) %>% 
  mutate(HaulPresence = ifelse(is.na(HaulPresence), 0, HaulPresence)) %>% 
  mutate(HaulProportion = ifelse(is.na(HaulProportion), 0, HaulProportion))



data_full$HaulBin <- ifelse(data_full$HaulProportion == 0.0, "Absent",
                        ifelse(data_full$HaulProportion > 0.0 & data_full$HaulProportion <= 0.10, "Rare",
                               ifelse(data_full$HaulProportion > 0.10 & data_full$HaulProportion <= 0.25, "Less Common",
                                      ifelse(data_full$HaulProportion > 0.25 & data_full$HaulProportion <= 0.75, "Common",
                                             ifelse(data_full$HaulProportion > 0.75 & data_full$HaulProportion <= 0.90, "Abundant",
                                                    ifelse(data_full$HaulProportion > 0.90, "Highly Abundant", NA))))))

data_full$HaulBin <- factor(data_full$HaulBin, levels = c("Absent", "Rare", "Less Common", "Common", "Abundant", "Highly Abundant"))
# saveRDS(data_full, file = "data_full.RData")

ggplot(data_full, aes(HaulBin)) +
  geom_bar( fill = "steelblue")

```

### Data Export
```{r exporting data_full}
# writing the csv for DisMAP development team here
data_export <- data_full %>% 
  select(Region, SurveyName, Year, Species, Common, Bin, WTCPUE, HaulProportion, HaulBin)

write.csv(data_export, here::here("data_processing_rcode","output","data_clean","SpeciesPersistenceIndicatorPercentileBin.csv"))

```


## Vizualizations

```{r}
#load packages needed for visualization section
library(tidyverse)
# library(hrbrthemes)
library(viridis)
library(plotly)
library(heatmaply)
```

The goal of this data processing is for users to be able to clearly see trends in the prevalence of species in surveys over time so that they can notice species that might be emerging or disappearing in a region.

**Note: Many of these visualizations are interactive. To see the values depicted in each cell, simply hover your cursor over the plot. To zoom in on a set of species or years, hold and drag your cursor to draw a box around the desired region.**

### All Surveys - Facet Wrapped
This code block needs to be changed to `eval = TRUE` to run/knit. Default is currently `eval = FALSE` because this chunk takes some time to process.
```{r all surveys viz, eval=FALSE, message=FALSE, warning=FALSE}

#### This takes a long time to knit so be prepared! ####

data_all <- data_full %>%
  ungroup()

surveys <- unique(data_all$SurveyName)

plotlist <- list()

for(v in surveys) {
  
  dat <- data_all %>% 
    filter(SurveyName == v)
  
  dat$Bin <- factor(dat$Bin, levels = c('Absent', 'Rare', 'Less Common', 'Common', 'Abundant', 'Highly Abundant'))
  
  p <- ggplot(dat, aes(Year, Common)) +
    geom_tile(aes(fill= Bin)) +
    ggtitle(paste(v)) +
    theme_minimal() +
    scale_fill_viridis(discrete=TRUE, option="viridis")
    # theme(axis.text.y = element_text(size = 8))
  
  # Convert ggplot to ggplotly and display
  plotlist[[v]] <- ggplotly(p)
  
}

htmltools::tagList(setNames(plotlist, NULL))

```


## Trend assessment

This section:
- 1) evaluates the trend categories based on observation (presence versus absence) in survey catch,
- 2) measures increasing versus decreasing trends in levels of haul frequency and biomass,
- 3) conducts a ground-truthing exercise to examine trends of a set of focal species,
- 4) applies a flagging function to trends that require further review, and
- 5) compiles all information for data export


### Trend Category Assignment

**Trend Category Definitions**

- *Persistently Present*: Consistently observed in surveys across years;
- *Established*: Species previously wasn't observed in surveys and now is consistently observed in surveys;
- *Potentially Emerging*: Species previously wasn't consistently observed in surveys but is potentially shifting towards being consistently observed;
- *Potentially Disappearing*: Species was previously observed in surveys but is potentially shifting towards being consistently not observed;
- *Disappeared*: Species was previously observed in surveys but now is consistently not observed in surveys;
- *Sporadic - Primarily Present*: Species is observed in over half of surveys but observations are inconsistent; and
- *Sporadic - Primarily Absent*: Species is observed in less than half of surveys and observations are inconsistent.

```{r organization}
data_cumm <- data_full %>% 
  select(Region, SurveyName, Year, Species, Common, Absence_Presence)

data_cumm$Absence_Presence <- as.numeric(data_cumm$Absence_Presence)
```

To do this, first we identify instances in which a species shifts from being present in a survey one year to being absent in a survey the following year, and vice versa.
```{r col vs ext}
holding <- data.frame(SurveyName = character(), Species = character(), Year = numeric(), Diff = numeric(), stringsAsFactors = FALSE)


surveys <- unique(data_cumm$SurveyName)

for(v in surveys) {

  firstfilt <- data_cumm %>%
    filter(SurveyName == v)

  spps <- unique(firstfilt$Species)

  for (p in spps) {

    myfilt <- firstfilt %>%
      filter(Species == p)

    years <- c(myfilt$Year)
    years <- years[order(years)]


    for (i in 1:(length(years)-1)) {

      var_1 <- myfilt %>%
        filter(Year == years[i])

      var_2 <- myfilt %>%
        filter(Year == years[i+1])

      Diff <- var_2$Absence_Presence[1] - var_1$Absence_Presence[1]

      temp <- data.frame(SurveyName = v, Species = p, Year = years[i+1], Diff = Diff, stringsAsFactors = FALSE)

      holding <- rbind(holding, temp)

    }
  }
}

```


```{r adding common names}
# Creating a species list will allow for joining in the common names at the end of the next chunk
spp_list <- distinct(data_full %>% 
  ungroup() %>% 
  select(SurveyName, Species, Common))

```


```{r}
# Identifying when species goes from absent to present in survey
data_ce_col <- holding %>% 
  filter(Diff == 1) %>% 
  group_by(SurveyName, Species) %>% 
  summarise(col = n())

# Identifying when species goes from present to absent in survey
data_ce_ext <- holding %>% 
  filter(Diff == -1) %>% 
  group_by(SurveyName, Species) %>% 
  summarise(ext = n())

# Identifying when species remains as either absent or present in survey
data_ce_stable <- holding %>% 
  filter(Diff == 0) %>% 
  group_by(SurveyName, Species) %>% 
  summarise(stable = n())

# merging three objects above
data_ce <- data_ce_stable %>%
  full_join(data_ce_col, by = c("SurveyName", "Species")) %>% 
  full_join(data_ce_ext, by = c("SurveyName", "Species")) %>% 
  replace(is.na(.), 0) %>% 
  select(-stable) %>% 
  left_join(spp_list, by = c("SurveyName", "Species")) %>% 
  select(SurveyName, Species, Common, col, ext)
```

```{r build data df}
#adding colonization and extinction event columns to data frame to use in trend categorization
data <- data_full %>% 
  left_join(data_ce, by = c("SurveyName", "Species", "Common"))

```

```{r function add_trend_category}

add_trend_category <- function(data){
    data %>%
    arrange(Region, SurveyName, Species, Common, Year) %>%
    group_by(Region, SurveyName, Species, Common) %>%
    group_modify(~{
      df <- .x
      n_years <- nrow(df)
      abs_vec <- as.numeric(df$Absence_Presence)
      
      # 1. Basic Stats
      percent_present <- mean(abs_vec, na.rm = TRUE)
      
      # 2. Windowed Stats (First 1/3 vs Last 1/3)
      third_n <- max(1, floor(n_years / 3))
      first_third <- abs_vec[1:third_n]
      last_third  <- abs_vec[(n_years - third_n + 1):n_years]
      
      mean_first <- mean(first_third, na.rm = TRUE)
      mean_last  <- mean(last_third, na.rm = TRUE)
      
      # 3. Recent Window (Last 5 years)
      last5_idx <- (max(1, n_years - 4)):n_years
      recent_5  <- abs_vec[last5_idx]
      sum_recent_5 <- sum(recent_5, na.rm = TRUE)
      
      # 4. New logic for the last 10 years
      last10_n <- min(10, n_years)
      last10_idx <- (n_years - last10_n + 1):n_years
      recent_10 <- abs_vec[last10_idx]
      mean_recent_10 <- mean(recent_10, na.rm = TRUE)
      
      # 3. Recent Window (Last 10 years)
      last10_idx <- (max(1, n_years - 9)):n_years
      recent_10  <- abs_vec[last10_idx]
      sum_recent_10 <- sum(recent_10, na.rm = TRUE)
      
      # 5. Run Length Encoding (Streaks)
      # Calculates the length of the final 'streak' of 1s or 0s
      runs <- rle(abs_vec)
      last_run_val <- tail(runs$values, 1)
      last_run_len <- tail(runs$lengths, 1)
      
      # 5. Logic Hierarchy (Established vs. Trending)
      trend_category <- case_when(
        
        # 1. PERSISTENTLY PRESENT
        # Species that were ALWAYS high (not a new establishment)
        # and any absences are not a recent streak of 3+ absences
        percent_present >= 0.85 & !(last_run_val == 0 & last_run_len >= 3)
        ~"Persistently Present",
          
        # 2, ABSENT / VAGRANT
        # It was very rarely seen (percent_present <= 0.05)
        # and if caught the hits aren't a recent streak of presences 
        percent_present <= 0.05 & !(last_run_val == 1 & last_run_len >= 3) ~ "Absent",
        
        # 3. ESTABLISHED / SUSTAINED (Emerged + Stabilized)
        # Low start/Sporadic start that is now high and consistent
        #A species that has "stabilized" usually satisfies two conditions:
          #High recent occupancy: They are present in almost all of the most recent years
          #Low early occupancy: They were infrequent or absent in the beginning of the time series.
        (mean_first <= 0.50 & mean_last >= 0.85 & last_run_val == 1 & last_run_len >=5) ~ "Established",
        
        # 4. POTENTIALLY EMERGING
        # Trend is upward, caught recently, but not 'solid' yet
        (mean_first <= 0.50 & mean_first < mean_last & sum_recent_5 >= 4) ~ "Potentially Emerging",
        
        # 5. DISSAPPEARED / SUSTAINED ABSENCE 
        # high start, but gone now
        (mean_first >= 0.50 & mean_last <= 0.15 & last_run_val == 0 & last_run_len >=5)
        ~ "Dissappeared",
        
        # 6. POTENTIALLY DISAPPEARING
        (mean_first >= 0.50 & mean_first > mean_last & sum_recent_5 <= 1) ~ "Potentially Disappearing",
        
        # 7. SPORADIC
        TRUE ~ "Sporadic"
      )
      
      df$TrendCategory <- trend_category
      return(df)
    }) %>%
    ungroup()
}

# Usage:
# This data frame will be used below in the flagging function
data_full_trends <- add_trend_category(data) %>%
  mutate(Name = paste(Common, SurveyName, sep = "_"))

# This dataframe will be used for the final data exporting
data_summary <- add_trend_category(data) %>%
  select(Region, SurveyName, Species, Common, TrendCategory) %>%
  distinct() %>% 
  mutate(SurveyName = as.character(SurveyName),
         Species = as.character(Species))
```


### Increasing or Decreasing Trend

This code aims to provide a similar analysis of that above, but with additional resolution. As opposed to solely focusing on presence/absence, here we add a second term that takes into account the abundance and/or haul frequency levels of each taxa over time to categorize each present trend as increasing or decreasing.

After evaluating the increase/decrease, this information is joined into the `data_summary` object, created above.


***Haul frequency***

This section is calculating a linear regression for the last 10 years to determine if there is any significant change (increase vs decrease) in *haul frequency*.
```{r organization haul freq}
data_inc_haul <- data_full %>% 
  select(Region, SurveyName, Year, Species, Common, HaulProportion)
```

To run a glm(), I need to break down the haul frequency term by breaking it back into haul present/total hauls
```{r}
haulsyr_max <- haulsyr_max %>%
  rename(
    SurveyName = survey,
    Year = year
  )
    
data_haul_glm <- data_full %>% 
  select(Region, SurveyName, Year, Species, Common, HaulPresence) %>% 
  left_join(haulsyr_max, by = c("SurveyName", "Year"))
```


```{r  haul freq glm, eval= TRUE}
holding_glm <- data.frame(SurveyName = character(), Species = character(), Year = numeric(), glmM = numeric(), glmP = numeric(), stringsAsFactors = FALSE)


surveys <- unique(data_haul_glm$SurveyName)

for(v in surveys) {

  firstfilt <- data_haul_glm %>%
    filter(SurveyName == v)

  spps <- unique(firstfilt$Species)

  for (p in spps) {

    myfilt <- firstfilt %>%
      filter(Species == p) %>%
      mutate(
        Year = as.numeric(Year), # character -> numeric 
        maxhauls = as.numeric(maxhauls) # safe conversion for response
        ) %>%
      filter(!is.na(Year), !is.na(HaulPresence), !is.na(maxhauls))
    
    # default output if regression is not possible
    glmM <- NA_real_
    glmP <- NA_real_
    YearVal <- NA_real_
    
    # determine the last-10-year window (min = max(Year) - 10, inclusive)
    if (nrow(myfilt) > 0 && any(!is.na(myfilt$Year))) {
      max_year <- max(myfilt$Year, na.rm = TRUE)
      min_year <- max_year - 10
      # subset to the last 10-year window (inclusive)
      window_filt <- myfilt %>% filter(Year >= min_year & Year <= max_year)
      } else {
      window_filt <- myfilt[0, ]  # empty
      }
    
    # Utilize `tryCatch()` to incorporate safe guards 
    if (nrow(window_filt) >= 2 && 
        dplyr::n_distinct(window_filt$Year) >= 2 &&
        sum(window_filt$HaulPresence, na.rm = TRUE) > 0 &&
        sum(window_filt$maxhauls - window_filt$HaulPresence, na.rm = TRUE) > 0) {
      
        fit_glm <- tryCatch({
          # !This section inside the brackets is the meat of the regression! 
          # To run this regression without the `tryCatch()` safeguards, the glm could be run on its own, but the `control  glm.control(maxit = 50)` could be removed from the function 
          glm(
          cbind(HaulPresence, maxhauls - HaulPresence) ~ Year,
          data = window_filt,
          family = binomial,
          control = glm.control(maxit = 50) # increase iterations if needed
          )
        }, warning = function(w) {
        # return the model anyway when a warning occurs to inspect it, or return NULL if you prefer
        # message("glm warning: ", conditionMessage(w))
        tryCatch(glm(
          cbind(HaulPresence, maxhauls - HaulPresence) ~ Year,
          data = window_filt,
          family = binomial,
          control = glm.control(maxit = 200)
        ), error = function(e) NULL)
        }, error = function(e) {
        # fit failed
        NULL
      })
        
        summary(fit_glm)
        
        
         if (!is.null(fit_glm)) {
          # use broom to extract coefficients and model-level stats, but guard with tryCatch
           tid <- tryCatch(broom::tidy(fit_glm), error = function(e) NULL)
           # gl  <- tryCatch(broom::glance(fit_glm), error = function(e) NULL)
          
           #extract slope and P-value (if present) using tidy output
           #if_else statements are used here to ensure that nothing is extracted from any tests that run as NULL
           if (!is.null(tid) && "term" %in% names(tid) && any(tid$term == "Year")) {
             yr_row <- tid[tid$term == "Year", , drop = FALSE]
             glmM <- if (!is.null(yr_row$estimate)) as.numeric(yr_row$estimate) else NA_real_
             glmP <- if ("p.value" %in% names(yr_row)) as.numeric(yr_row$p.value) else NA_real_
           }
           
          YearVal <- max(window_filt$Year, na.rm = TRUE)#FLAG:This is likely not needed as we drop this column later
         }
    }
    
    temp <- data.frame(
      SurveyName = v,
      Species = p,
      Year = YearVal,
      glmM = glmM,
      glmP = glmP,
      stringsAsFactors = FALSE
    )

    holding_glm <- rbind(holding_glm, temp)
  }
}
  
```

```{r filter sig haul trend}
holding_noyear <- holding_glm %>% select(!Year)

holding_noyear_filt <- holding_glm %>% select(!Year) %>% 
  filter(glmP <= 0.05)

holding_noyear_filt$Haul_Inc_Dec <- ifelse(holding_noyear_filt$glmM > 0, "Increasing", 
                                      ifelse(holding_noyear_filt$glmM < 0, "Decreasing", NA))
```


***Biomass***

This section is calculating a linear regression for the last 10 years to determine if there is any significant change (increase vs decrease) in *biomass*.
```{r organization biomass}
data_inc_wtcpue <- data_full %>% 
  select(Region, SurveyName, Year, Species, Common, WTCPUE)
```

```{r biomass inc vs dec}
holding_bio <- data.frame(SurveyName = character(), Species = character(), Year = numeric(), RegM = numeric(), RegR = numeric(), RegP = numeric(), stringsAsFactors = FALSE)


surveys <- unique(data_inc_wtcpue$SurveyName)

for(v in surveys) {

  firstfilt <- data_inc_wtcpue %>%
    filter(SurveyName == v)

  spps <- unique(firstfilt$Species)

  for (p in spps) {

    myfilt <- firstfilt %>%
      filter(Species == p) %>%
      mutate(
        Year = as.numeric(Year), # character -> numeric 
        WTCPUE = as.numeric(as.character(WTCPUE)) # safe conversion for response
        ) %>%
      filter(!is.na(Year), !is.na(WTCPUE))
    
    myfilt$Year <- as.numeric(myfilt$Year)
    
    # default output if regression is not possible
    RegM <- NA_real_
    RegR <- NA_real_
    RegP <- NA_real_
    YearVal <- NA_real_
    
    # determine the last-10-year window (min = max(Year) - 10, inclusive)
    if (nrow(myfilt) > 0 && any(!is.na(myfilt$Year))) {
      max_year <- max(myfilt$Year, na.rm = TRUE)
      min_year <- max_year - 10
      # subset to the last 10-year window (inclusive)
      window_filt <- myfilt %>% filter(Year >= min_year & Year <= max_year)
      } else {
      window_filt <- myfilt[0, ]  # empty
      }
    
    if (nrow(window_filt) >= 2 && 
        dplyr::n_distinct(window_filt$Year) >= 2 &&
        sd(window_filt$WTCPUE, na.rm = TRUE) > 0) {
        
        model <- tryCatch(
          lm(WTCPUE ~ Year, data = window_filt),
          error = function(e) NULL
          )
        
         if (!is.null(model)) {
          # use broom to extract coefficients and model-level stats, but guard with tryCatch
           tid <- tryCatch(broom::tidy(model), error = function(e) NULL)
           gl  <- tryCatch(broom::glance(model), error = function(e) NULL)
          
           #extract slope and P-value (if present) using tidy output
           #if_else statements are used here to ensure that nothing is extracted from any tests that run as NULL
           if (!is.null(tid) && "term" %in% names(tid) && any(tid$term == "Year")) {
             yr_row <- tid[tid$term == "Year", , drop = FALSE]
             RegM <- if (!is.null(yr_row$estimate)) as.numeric(yr_row$estimate) else NA_real_
             RegP <- if ("p.value" %in% names(yr_row)) as.numeric(yr_row$p.value) else NA_real_
           }
           
           #extract adjusted R-squared value using glance output
           if (!is.null(gl) && "adj.r.squared" %in% names(gl)) {
             RegR <- as.numeric(gl$adj.r.squared)
           } else{
             RegR<- NA_real_
           }
           
          YearVal <- max(window_filt$Year, na.rm = TRUE)#FLAG:This is likely not needed as we drop this column later
         }
    }
    
    temp <- data.frame(
      SurveyName = v,
      Species = p,
      Year = YearVal,
      RegM = RegM,
      RegR = RegR,
      RegP= RegP,
      stringsAsFactors = FALSE
    )

    holding_bio <- rbind(holding_bio, temp)
  }
}

```

```{r filter sig biomass trend}
holding_bio_noyear <- holding_bio %>% select(!Year)

holding_bio_noyear_filt <- holding_bio %>% select(!Year) %>% 
  filter(RegP <= 0.05)

holding_bio_noyear_filt$Bio_Inc_Dec <- ifelse(holding_bio_noyear_filt$RegM > 0, "Increasing", 
                                      ifelse(holding_bio_noyear_filt$RegM < 0, "Decreasing", NA))
```

Add these columns into the `data_summary` object above
```{r compile with data_summary}

data_summary <- data_summary %>% 
  left_join(holding_noyear_filt, by = c("SurveyName", "Species")) %>% 
  select(-glmM, -glmP) %>% 
  left_join(holding_bio_noyear_filt, by = c("SurveyName", "Species")) %>% 
  select(-RegM, -RegR, -RegP)
```


### Ground-truthing

This section visualizes the trends in *haul frequency* and *biomass* for a set of "focal species". The goal of these visualizations is to help us ground-truth our analyses and determine if the trends of these "focal species" successfully match the trends that we know they should have in each region. The last 10 years are represented by a gray window to highlight that these years are captured by the linear regression model.

We also provide a subset of the visualizations that only include "focal species" with significant trends (p<=0.5).

```{r haul freq}

focal_spp <- data_inc_haul %>% 
  filter((Region== "Aleutian Islands" & Common == "Walleye pollock")|
        (Region== "Eastern Bering Sea" & Common == "Walleye pollock")|
        (Region== "Gulf of Alaska" & Common == "Walleye pollock")|
        (Region== "Northern Bering Sea" & Common == "Walleye pollock")|
        (Region== "Aleutian Islands" & Common == "Sablefish")|
        (Region== "Eastern Bering Sea" & Common == "Sablefish")|
        (Region== "Gulf of Alaska" & Common == "Sablefish")|
        (Region== "Northern Bering Sea" & Common == "Sablefish")|
        (Region== "Aleutian Islands" & Common == "Pacific cod")|
        (Region== "Eastern Bering Sea" & Common == "Pacific cod")|
        (Region== "Gulf of Alaska" & Common == "Pacific cod")|
        (Region== "Northern Bering Sea" & Common == "Pacific cod")|
        (Region== "Gulf of Mexico" & Common == "Red snapper")|
        (Region== "Gulf of Mexico" & Common == "Gray snapper")|
        (Region== "Northeast US" & Common == "Atlantic northern shrimp")|
        (Region== "Northeast US" & Common == "Atlantic cod")|
        (Region== "Northeast US" & Common == "American lobster")|
        (Region== "Northeast US" & Common == "Black sea bass")|
        (Region== "Southeast US" & Common == "Pink shrimp")|
        (Region== "Southeast US" & Common == "King mackerel")|
        (Region== "West Coast" & Common == "Yelloweye rockfish")|
        (Region== "West Coast" & Common == "Pacific halibut")|
        (Region== "West Coast" & Common == "Sablefish") |
        (Region== "Hawai'i Islands" & Common =="Hapu'upu'u")|
        (Region== "Hawai'i Islands" & Common == "Gindai")| 
        (Region== "Hawai'i Islands" & Common =="Lehi")|
        (Region== "Hawai'i Islands" & Common == "Ehu")|
        (Region== "Hawai'i Islands" & Common == "Onaga")| 
        (Region== "Hawai'i Islands" & Common =="Opakapaka")|
        (Region== "Hawai'i Islands" & Common == "Kalekale")
          ) %>% 
  left_join(holding_noyear, by=c("SurveyName", "Species")) %>%
  mutate(id = paste(Common, abbreviate(SurveyName, 11), sep = ", "))

focal_spp$Year <- as.numeric(focal_spp$Year)

#same as above but filtering for only significant p-values
  
focal_spp_filt <- data_inc_haul %>% 
  filter((Region== "Aleutian Islands" & Common == "Walleye pollock")|
        (Region== "Eastern Bering Sea" & Common == "Walleye pollock")|
        (Region== "Gulf of Alaska" & Common == "Walleye pollock")|
        (Region== "Northern Bering Sea" & Common == "Walleye pollock")|
        (Region== "Aleutian Islands" & Common == "Sablefish")|
        (Region== "Eastern Bering Sea" & Common == "Sablefish")|
        (Region== "Gulf of Alaska" & Common == "Sablefish")|
        (Region== "Northern Bering Sea" & Common == "Sablefish")|
        (Region== "Aleutian Islands" & Common == "Pacific cod")|
        (Region== "Eastern Bering Sea" & Common == "Pacific cod")|
        (Region== "Gulf of Alaska" & Common == "Pacific cod")|
        (Region== "Northern Bering Sea" & Common == "Pacific cod")|
        (Region== "Gulf of Mexico" & Common == "Red snapper")|
        (Region== "Gulf of Mexico" & Common == "Gray snapper")|
        (Region== "Northeast US" & Common == "Atlantic northern shrimp")|
        (Region== "Northeast US" & Common == "Atlantic cod")|
        (Region== "Northeast US" & Common == "American lobster")|
        (Region== "Northeast US" & Common == "Black sea bass")|
        (Region== "Southeast US" & Common == "Pink shrimp")|
        (Region== "Southeast US" & Common == "King mackerel")|
        (Region== "West Coast" & Common == "Yelloweye rockfish")|
        (Region== "West Coast" & Common == "Pacific halibut")|
        (Region== "West Coast" & Common == "Sablefish")|
        (Region== "Hawai'i Islands" & Common =="Hapu'upu'u")|
        (Region== "Hawai'i Islands" & Common == "Gindai")| 
        (Region== "Hawai'i Islands" & Common =="Lehi")|
        (Region== "Hawai'i Islands" & Common == "Ehu")|
        (Region== "Hawai'i Islands" & Common == "Onaga")| 
        (Region== "Hawai'i Islands" & Common =="Opakapaka")|
        (Region== "Hawai'i Islands" & Common == "Kalekale")
          ) %>% 
  left_join(holding_noyear_filt, by=c("SurveyName", "Species")) %>%
  mutate(id = paste(Common, abbreviate(SurveyName, 11), sep = ", ")) %>% 
  filter(!is.na(RegP))

focal_spp_filt$Year <- as.numeric(focal_spp_filt$Year)

```

```{r viz 1}
library(plotly)

focal_check_haul <- ggplot(focal_spp, aes(x = Year, y = HaulProportion)) +
  annotate("rect", xmin= 2014, xmax = 2024, ymin = 0.00, ymax= 1.00, alpha = 0.2) +
  geom_point(aes(color= c(id))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Haul Freq. - Full Focal Spp")

ggplotly(focal_check_haul)
```


```{r viz 2}
focal_check_haul_filt <- ggplot(focal_spp_filt, aes(x = Year, y = HaulProportion)) +
  annotate("rect", xmin= 2014, xmax = 2024, ymin = 0.00, ymax= 1.00, alpha = 0.2) +
  geom_point(aes(color= c(id))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Haul Freq. - Focal Spp w/ p<=0.05")

ggplotly(focal_check_haul_filt)
```


```{r wtcpue}
focal_spp_bio <- data_inc_wtcpue %>% 
  filter((Region== "Aleutian Islands" & Common == "Walleye pollock")|
        (Region== "Eastern Bering Sea" & Common == "Walleye pollock")|
        (Region== "Gulf of Alaska" & Common == "Walleye pollock")|
        (Region== "Northern Bering Sea" & Common == "Walleye pollock")|
        (Region== "Aleutian Islands" & Common == "Sablefish")|
        (Region== "Eastern Bering Sea" & Common == "Sablefish")|
        (Region== "Gulf of Alaska" & Common == "Sablefish")|
        (Region== "Northern Bering Sea" & Common == "Sablefish")|
        (Region== "Aleutian Islands" & Common == "Pacific cod")|
        (Region== "Eastern Bering Sea" & Common == "Pacific cod")|
        (Region== "Gulf of Alaska" & Common == "Pacific cod")|
        (Region== "Northern Bering Sea" & Common == "Pacific cod")|
        (Region== "Gulf of Mexico" & Common == "Red snapper")|
        (Region== "Gulf of Mexico" & Common == "Gray snapper")|
        (Region== "Northeast US" & Common == "Atlantic northern shrimp")|
        (Region== "Northeast US" & Common == "Atlantic cod")|
        (Region== "Northeast US" & Common == "American lobster")|
        (Region== "Northeast US" & Common == "Black sea bass")|
        (Region== "Southeast US" & Common == "Pink shrimp")|
        (Region== "Southeast US" & Common == "King mackerel")|
        (Region== "West Coast" & Common == "Yelloweye rockfish")|
        (Region== "West Coast" & Common == "Pacific halibut")|
        (Region== "West Coast" & Common == "Sablefish")|
        (Region== "Hawai'i Islands" & Common =="Hapu'upu'u")|
        (Region== "Hawai'i Islands" & Common == "Gindai")| 
        (Region== "Hawai'i Islands" & Common =="Lehi")|
        (Region== "Hawai'i Islands" & Common == "Ehu")|
        (Region== "Hawai'i Islands" & Common == "Onaga")| 
        (Region== "Hawai'i Islands" & Common =="Opakapaka")|
        (Region== "Hawai'i Islands" & Common == "Kalekale")
          ) %>% 
  left_join(holding_bio_noyear, by=c("SurveyName", "Species")) %>% 
  mutate(id = paste(Common, abbreviate(SurveyName, 11), sep = ", "))

focal_spp_bio$Year <- as.numeric(focal_spp_bio$Year)

#filtering for significant p-values
focal_spp_bio_filt <- data_inc_wtcpue %>% 
  filter((Region== "Aleutian Islands" & Common == "Walleye pollock")|
        (Region== "Eastern Bering Sea" & Common == "Walleye pollock")|
        (Region== "Gulf of Alaska" & Common == "Walleye pollock")|
        (Region== "Northern Bering Sea" & Common == "Walleye pollock")|
        (Region== "Aleutian Islands" & Common == "Sablefish")|
        (Region== "Eastern Bering Sea" & Common == "Sablefish")|
        (Region== "Gulf of Alaska" & Common == "Sablefish")|
        (Region== "Northern Bering Sea" & Common == "Sablefish")|
        (Region== "Aleutian Islands" & Common == "Pacific cod")|
        (Region== "Eastern Bering Sea" & Common == "Pacific cod")|
        (Region== "Gulf of Alaska" & Common == "Pacific cod")|
        (Region== "Northern Bering Sea" & Common == "Pacific cod")|
        (Region== "Gulf of Mexico" & Common == "Red snapper")|
        (Region== "Gulf of Mexico" & Common == "Gray snapper")|
        (Region== "Northeast US" & Common == "Atlantic northern shrimp")|
        (Region== "Northeast US" & Common == "Atlantic cod")|
        (Region== "Northeast US" & Common == "American lobster")|
        (Region== "Northeast US" & Common == "Black sea bass")|
        (Region== "Southeast US" & Common == "Pink shrimp")|
        (Region== "Southeast US" & Common == "King mackerel")|
        (Region== "West Coast" & Common == "Yelloweye rockfish")|
        (Region== "West Coast" & Common == "Pacific halibut")|
        (Region== "West Coast" & Common == "Sablefish")|
        (Region== "Hawai'i Islands" & Common =="Hapu'upu'u")|
        (Region== "Hawai'i Islands" & Common == "Gindai")| 
        (Region== "Hawai'i Islands" & Common =="Lehi")|
        (Region== "Hawai'i Islands" & Common == "Ehu")|
        (Region== "Hawai'i Islands" & Common == "Onaga")| 
        (Region== "Hawai'i Islands" & Common =="Opakapaka")|
        (Region== "Hawai'i Islands" & Common == "Kalekale")
          ) %>% 
  left_join(holding_bio_noyear_filt, by=c("SurveyName", "Species")) %>% 
  mutate(id = paste(Common, abbreviate(SurveyName, 11), sep = ", "))%>% 
  filter(!is.na(RegP))

focal_spp_bio_filt$Year <- as.numeric(focal_spp_bio_filt$Year)

```

```{r viz 3}
focal_check_bio <- ggplot(focal_spp_bio, aes(x = Year, y = WTCPUE)) +
  annotate("rect", xmin= 2014, xmax = 2024, ymin = 0, ymax= 65000, alpha = 0.2) +
  geom_point(aes(color= c(id))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("WTCPUE - Full Focal Spp.")

ggplotly(focal_check_bio)
```

```{r viz 4}
focal_check_bio_filt <- ggplot(focal_spp_bio_filt, aes(x = Year, y = WTCPUE)) +
  annotate("rect", xmin= 2014, xmax = 2024, ymin = 0, ymax= 65000, alpha = 0.2) +
  geom_point(aes(color= c(id))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  ggtitle("WTCPUE - Focal Spp w/ p<=0.05")

ggplotly(focal_check_bio_filt)
```

```{r non significant haul removed, eval=FALSE}

focal_spp_haulremoved <- data_inc_haul %>% 
  filter((Region== "Aleutian Islands" & Common == "Walleye pollock")|
        (Region== "Eastern Bering Sea" & Common == "Walleye pollock")|
        (Region== "Gulf of Alaska" & Common == "Walleye pollock")|
        (Region== "Northern Bering Sea" & Common == "Walleye pollock")|
        (Region== "Aleutian Islands" & Common == "Sablefish")|
        (Region== "Eastern Bering Sea" & Common == "Sablefish")|
        (Region== "Gulf of Alaska" & Common == "Sablefish")|
        (Region== "Northern Bering Sea" & Common == "Sablefish")|
        (Region== "Aleutian Islands" & Common == "Pacific cod")|
        (Region== "Eastern Bering Sea" & Common == "Pacific cod")|
        (Region== "Gulf of Alaska" & Common == "Pacific cod")|
        (Region== "Northern Bering Sea" & Common == "Pacific cod")|
        (Region== "Gulf of Mexico" & Common == "Red snapper")|
        (Region== "Gulf of Mexico" & Common == "Gray snapper")|
        (Region== "Northeast US" & Common == "Atlantic northern shrimp")|
        (Region== "Northeast US" & Common == "Atlantic cod")|
        (Region== "Northeast US" & Common == "American lobster")|
        (Region== "Northeast US" & Common == "Black sea bass")|
        (Region== "Southeast US" & Common == "Pink shrimp")|
        (Region== "Southeast US" & Common == "King mackerel")|
        (Region== "West Coast" & Common == "Yelloweye rockfish")|
        (Region== "West Coast" & Common == "Pacific halibut")|
        (Region== "West Coast" & Common == "Sablefish")
          ) %>% 
  left_join(holding_noyear_filt, by=c("SurveyName", "Species")) %>%
  mutate(id = paste(Common, abbreviate(SurveyName, 11), sep = ", ")) %>% 
  filter(is.na(RegP))

focal_spp_haulremoved$Year <- as.numeric(focal_spp_haulremoved$Year)

focal_check_haul_removed <- ggplot(focal_spp_haulremoved, aes(x = Year, y = HaulProportion)) +
  annotate("rect", xmin= 2014, xmax = 2024, ymin = 0.00, ymax= 1.00, alpha = 0.2) +
  geom_point(aes(color= c(id))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Haul Freq - REMOVED Focal, p>0.05")

ggplotly(focal_check_haul_removed)
```


```{r non significant bio removed, eval=FALSE}

focal_spp_bioremoved <- data_inc_wtcpue %>% 
  filter((Region== "Aleutian Islands" & Common == "Walleye pollock")|
        (Region== "Eastern Bering Sea" & Common == "Walleye pollock")|
        (Region== "Gulf of Alaska" & Common == "Walleye pollock")|
        (Region== "Northern Bering Sea" & Common == "Walleye pollock")|
        (Region== "Aleutian Islands" & Common == "Sablefish")|
        (Region== "Eastern Bering Sea" & Common == "Sablefish")|
        (Region== "Gulf of Alaska" & Common == "Sablefish")|
        (Region== "Northern Bering Sea" & Common == "Sablefish")|
        (Region== "Aleutian Islands" & Common == "Pacific cod")|
        (Region== "Eastern Bering Sea" & Common == "Pacific cod")|
        (Region== "Gulf of Alaska" & Common == "Pacific cod")|
        (Region== "Northern Bering Sea" & Common == "Pacific cod")|
        (Region== "Gulf of Mexico" & Common == "Red snapper")|
        (Region== "Gulf of Mexico" & Common == "Gray snapper")|
        (Region== "Northeast US" & Common == "Atlantic northern shrimp")|
        (Region== "Northeast US" & Common == "Atlantic cod")|
        (Region== "Northeast US" & Common == "American lobster")|
        (Region== "Northeast US" & Common == "Black sea bass")|
        (Region== "Southeast US" & Common == "Pink shrimp")|
        (Region== "Southeast US" & Common == "King mackerel")|
        (Region== "West Coast" & Common == "Yelloweye rockfish")|
        (Region== "West Coast" & Common == "Pacific halibut")|
        (Region== "West Coast" & Common == "Sablefish")
          ) %>% 
  left_join(holding_bio_noyear_filt, by=c("SurveyName", "Species")) %>% 
  mutate(id = paste(Common, abbreviate(SurveyName, 11), sep = ", "))%>% 
  filter(is.na(RegP))

focal_spp_bioremoved$Year <- as.numeric(focal_spp_bioremoved$Year)

focal_check_bio_removed <- ggplot(focal_spp_bioremoved, aes(x = Year, y = WTCPUE)) +
  annotate("rect", xmin= 2014, xmax = 2024, ymin = 0, ymax= 65000, alpha = 0.2) +
  geom_point(aes(color= c(id))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  ggtitle("WTCPUE - REMOVED Focal Spp, p>0.05")

ggplotly(focal_check_bio_removed)
```

```{r pvalue hist, eval = FALSE}
pval_haul <- ggplot(holding, aes(x = RegP)) +
  geom_histogram() +
  theme_minimal()

```


### Flagging function

The following two code chunks adds a flag for species trends that are under review. Currently, the flagging isn't being used, but the function below easily could be adjusted for future scenarios that require flagging. 

The current flagging function below will *not* capture all trends that require review and might incorrectly flag some trends that do not require review.

```{r species flag function, eval=TRUE}

data <- data_full_trends

flag_spp <- function(data, survey){
  
  cat("taxonomy check for survey: ", survey, sep="\n")
  
  #select region (survey column)
  data_survey <- data %>% 
    dplyr::mutate(SurveyName = forcats::as_factor(SurveyName)) %>% 
    dplyr::filter(SurveyName == survey)
  
  
  #get grouped species names per year
  df <- data_survey %>%
    dplyr::select(Species, Year, Absence_Presence) %>%
    dplyr::group_by(Species) %>%
    dplyr::distinct()

  
  #make species x year matrix
  years <- unique(df$Year)
  df2 <- data.frame(matrix(ncol = length(years)+1, nrow = length(df$Species)))
  x <- c('spp', unique(df$Year))
  colnames(df2) <- x
  df2$spp <- df$Species
  
  #fill up matrix with FALSE/TRUE depending on species presence/absence
  n <- 2
  for (y in years) {
    df2[, n] <- sapply(df2$spp, function(spp_name) {
      any(df$Species == spp_name & df$Year == y & df$Absence_Presence == 1)
      })
    n <- n + 1
    }
  
  #summarize
  df2 <- df2 %>%
    dplyr::group_by(spp) %>%
    dplyr::summarize_all(any)
  
  #create new dataframe
  df3 <- data.frame(matrix(ncol = 3, nrow = length(df2$spp)))
  colnames(df3) <- c('spp', 'rlmax', 'PresToAbs')
  
  #fill up the dataframe
  for (n in 1:length(df2$spp)){
    
    #Compute the lengths and values of runs of equal values in a vector
    r <- rle(df2[n,])
    
    sp <- c(df2[n,1])
    
    rlmax <- max(r$lengths)
    
    df3[n,1] <- sp
    df3[n,2] <- rlmax #maximum run length
    df3[n,3] <- length(r$values) #total nb of runs
    
  }
  
  #get species with maximum run length lower than 95% of the years (56)
  #and that occurred in less than 4 runs
  x <- round(0.95*length(years))
  df3 <- df3[df3$rlmax < x,]
  df3 <- df3[df3$PresToAbs < 4,]
  
  testdf <- data.frame(spp = df3$spp)
  
  if (dim(testdf)[1] > 0) {
    
    #write txt of flagged spp in region
    # sink(here::here("outputs", "Flags", "taxonomic_flagging", paste0(region, "_flagspp.txt")))
    # cat(testdf$spp, sep = " ; ")
    # sink()
    
    #write stats of species flagged
    # data_out <- data.frame("name" = c("Total number of species", "Percentage of species flagged"),
    #                    "nb" = c(length(df2$spp), round(100*(length(testdf$spp) /  length(df2$spp)), 1)))
    # write.csv(data_out, file = here::here("data_processing_rcode", "output", paste0(survey, "_stats.csv")), row.names = FALSE)
    
    #write df of species flagged
    survey_clean <- gsub("[/\\]", "_", survey)  # Replace slashes/backslashes with underscore
    write.csv(df3, file = here::here("data_processing_rcode", "output", "flags", paste0(survey_clean, "_flags.csv")), row.names = FALSE)
    
    #create absence dataframe for plotting
    allyears <- min(years):max(years)
    abs <- data.frame("accepted_name" = rep(df3$spp, each = length(allyears)),
                      "year" = rep(allyears, length(df3$spp)))
    
    #get presence dataframe for plotting
    pres <- subset(data_survey, Species %in% testdf$spp) %>%
      dplyr::select(Species, Year)

  }else{
    
    print("----- No species flagged for this region")
  }
}
```


```{r flag compilation, eval=TRUE}

survey_list <- unique(data_full_trends$SurveyName)

for (i in survey_list) {
  flag_spp(data, survey = i)
}


#bringing csv files back into combine them into one df
folder_path <- here::here("data_processing_rcode", "output", "flags")

files <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE)
# files_clean <- paste0(list.files(path = folder_path, pattern = "\\.csv$", full.names = FALSE)) 
  
df_list <- list()

for (n in seq_along(files)){
  #read in csv
  temp_df <- read_csv(files[n])
  # Get the file name without path and extension
  name <- tools::file_path_sans_ext(basename(files[n]))
  # Remove the string "flag" at the end, if present
  name <- sub("_flags$", "", name)
  # Add the source column as the first column
  temp_df <- cbind(survey = name, temp_df)
  # Store in list
  df_list[[n]] <- temp_df
}

flags_combo <- do.call(rbind, df_list) %>% 
  rename(SurveyName = survey,
         Species = spp) %>% 
  mutate(SurveyName = gsub("_", "/", SurveyName))

flags_combo$Notes <- c(rep("Preliminary trend - under review", nrow(flags_combo)))

```

### Data Export
```{r export data_summary}

data_trends_export <- data_summary %>% 
  left_join(flags_combo, by = c("SurveyName", "Species")) %>% 
  select(Region, SurveyName, Species, Common, TrendCategory, Notes, Haul_Inc_Dec, Bio_Inc_Dec)

write.csv(data_trends_export, here::here("data_processing_rcode","output","data_clean","SpeciesPersistenceIndicatorTrend.csv"))
```


The following table provides the number of species that fall into each category, using this version of the metrics.
```{r summary table}

library(knitr)

trendsummary <- data_summary %>%
  group_by(TrendCategory) %>%
  summarize(n = n())

kable(trendsummary)

```


```{r trend viz, eval = FALSE}
#Persistently Present 
trend_pres <- data_full_trends %>% 
  filter(TrendCategory == "Persistently Present")

ggplot(trend_pres, aes(Year, Name)) +
  geom_tile(aes(fill= Absence_Presence)) +
  ggtitle("Persistently Present") +
  theme_minimal() +
  scale_fill_viridis(discrete=TRUE, option="viridis") +
  scale_y_discrete(label = function(x) stringr::str_trunc(x, 15))

#Established
trend_estab <- data_full_trends %>% 
  filter(TrendCategory == "Established")

p6 <- ggplot(trend_estab, aes(Year, Name)) +
  geom_tile(aes(fill= Absence_Presence)) +
  ggtitle("Established") +
  theme_minimal() +
  scale_fill_viridis(discrete=TRUE, option="viridis") +
  scale_y_discrete(label = function(x) stringr::str_trunc(x, 15))

ggplotly(p6)


#Disappeared 
trend_disapp <- data_full_trends %>% 
  filter(TrendCategory == "Disappeared")

p2 <- ggplot(trend_disapp, aes(Year, Name)) +
  geom_tile(aes(fill= Absence_Presence)) +
  ggtitle("Disappeared") +
  theme_minimal() +
  scale_fill_viridis(discrete=TRUE, option="viridis") +
  scale_y_discrete(label = function(x) stringr::str_trunc(x, 15))

ggplotly(p2)

#Potentially Emerging 
trend_emerging <- data_full_trends %>% 
  filter(TrendCategory == "Potentially Emerging")

p3 <- ggplot(trend_emerging, aes(Year, Name)) +
  geom_tile(aes(fill= Absence_Presence)) +
  ggtitle("Potentially Emerging") +
  theme_minimal() +
  scale_fill_viridis(discrete=TRUE, option="viridis") +
  scale_y_discrete(label = function(x) stringr::str_trunc(x, 15))

ggplotly(p3)

#Potentially Disappearing 
trend_potdis <- data_full_trends %>% 
  filter(TrendCategory == "Potentially Disappearing")

p7 <- ggplot(trend_potdis, aes(Year, Name)) +
  geom_tile(aes(fill= Absence_Presence)) +
  ggtitle("Potentially Disappearing") +
  theme_minimal() +
  scale_fill_viridis(discrete=TRUE, option="viridis") +
  scale_y_discrete(label = function(x) stringr::str_trunc(x, 15))

ggplotly(p7)

#Sporadic 
trend_spopre <- data_full_trends %>% 
  filter(TrendCategory == "Sporadic")

p5 <- ggplot(trend_spopre, aes(Year, Name)) +
  geom_tile(aes(fill= Absence_Presence)) +
  ggtitle("Sporadic") +
  theme_minimal() +
  scale_fill_viridis(discrete=TRUE, option="viridis") +
  scale_y_discrete(label = function(x) stringr::str_trunc(x, 15))

ggplotly(p5)


```

##MANAGEMENT COUNCIL JURISDICTION ANALYSIS (for the NEFSC survey data only)
The following script does the same analysis as above, but parses the NE Bottom trawl data out into the two different management council jurisdictions. 
``` {r classify by fishery jurisdiction, eval = TRUE}

library(sf)
library(dplyr)

# 1. Load multi-polygon shapefile
boundaries <- st_read(here::here("data_processing_rcode/data/fishery_management_council_regions_20210609", "20210609_fishery_management_council_regions.shp"))
boundaries_exploded<-st_cast(boundaries, "POLYGON")
selected_boundaries<-boundaries_exploded %>%
  filter(FishRegion %in% c("New England", "NEMA", "South Atlantic")) 
# Apply the fix duplicate vertices issues
map_cleaned <- st_make_valid(selected_boundaries)
# Casts each individual Polygon back into its own Multipolygon wrapper
EC_Councils <- st_cast(map_cleaned, "MULTIPOLYGON")

# 2. Convert your dataframe to a spatial object
# We'll create a copy so we don't overwrite the original 'df' immediately
pts_sf <- st_as_sf(data_update, coords = c("lon", "lat"), crs = 4326, remove=FALSE)

# 3. Align the Coordinate Reference Systems (Crucial!)
EC_Councils <- st_transform(EC_Councils, st_crs(pts_sf))

# 4. Perform a Left Spatial Join
# 'left = TRUE' ensures we keep ALL points, even those that fall outside both polygons
df_labeled_sf <- st_join(pts_sf, EC_Councils, join = st_intersects, left = TRUE)
df_labeled_sf_NEUS <- df_labeled_sf %>%
  filter(region=="Northeast US")

# 5. Clean up and return to a standard dataframe
df_NEUS <- df_labeled_sf_NEUS %>%
  st_drop_geometry() %>%
  select(region, survey, haulid, year, spp, common, wtcpue, stratum, stratumarea, lat, lon, depth, FishRegion) %>%
  #filter(!is.na(FishRegion)) %>%
  unite("FishRegionsurvey", FishRegion, survey, sep=" ", remove=FALSE)

# 6. Visualize to make sure the points are labeled correctly 
ggplot() +
  geom_sf(data = EC_Councils, aes(fill = FishRegion), alpha = 0.3) +
  geom_point(data = df_NEUS, aes(x=lon, y=lat, color = FishRegion)) +
  theme_minimal() +
  labs(title = "Points Labeled by Fishery Management Council Jurisdiction")

```

#Proportion of total biomass by species in each Council region 
We calculate total biomass for each species within each Council region for each year by survey, and then calculate the total biomass for each species caught in the NEFSC trawl survey, and determine what proportion of that total biomass falls within each Council Region. 

```{r}
data_sum_NEUSregions <- df_NEUS %>%
  select(region, survey, year, spp, common, wtcpue, FishRegion, FishRegionsurvey) %>%
  group_by(region, survey, year, spp, common, FishRegion, FishRegionsurvey) %>%
  summarise(FMCwtcpue = sum(wtcpue))

data_sum_Total <- df_NEUS %>%
  select(region, survey, year, spp, common, wtcpue) %>%
  group_by(region, survey, year, spp, common) %>%
  summarise(Totwtcpue = sum(wtcpue))

data_sum_All <- left_join(data_sum_NEUSregions, data_sum_Total, by=c("region", "survey", "year", "spp", "common"))%>%
  filter(!is.na(FishRegion)) %>%
  mutate(prop=if_else(Totwtcpue ==0, 0, FMCwtcpue/Totwtcpue))

```

## Presistence trend categories by species, FMC region, and survey 

```{r}
#add a presenece vs. absense column to the dataframe 
data_sum_All$Absence_Presence <- ifelse(data_sum_All$FMCwtcpue == 0, "0",
                                     ifelse(data_sum_All$FMCwtcpue > 0, "1", "NA"))

data_cumm <- data_sum_All %>% 
  select(region, survey, year, spp, common, FishRegion, FishRegionsurvey, Absence_Presence)

data_cumm$Absence_Presence <- as.numeric(data_cumm$Absence_Presence)
```

To do this, first we identify instances in which a species shifts from being present in a FMC survey region one year to being absent in a survey the following year, and vice versa.
```{r colinization vs. extinction events}
holding <- data.frame(FishRegionsurvey = character(), spp = character(), year = numeric(), Diff = numeric(), stringsAsFactors = FALSE)


FMCsurveys <- unique(data_cumm$FishRegionsurvey)

for(v in FMCsurveys) {

  firstfilt <- data_cumm %>%
    filter(FishRegionsurvey == v)

  spps <- unique(firstfilt$spp)

  for (p in spps) {

    myfilt <- firstfilt %>%
      filter(spp == p)

    years <- c(myfilt$year)
    years <- years[order(years)]


    for (i in 1:(length(years)-1)) {

      var_1 <- myfilt %>%
        filter(year == years[i])

      var_2 <- myfilt %>%
        filter(year == years[i+1])

      Diff <- var_2$Absence_Presence[1] - var_1$Absence_Presence[1]

      temp <- data.frame(FishRegionsurvey = v, spp = p, year = years[i+1], Diff = Diff, stringsAsFactors = FALSE)

      holding <- rbind(holding, temp)

    }
  }
}

```


```{r adding common}
# Creating a species list will allow for joining in the common names at the end of the next chunk
spp_list <- distinct(data_sum_All %>% 
  ungroup() %>% 
  select(FishRegionsurvey, spp, common))

```


```{r}
# Identifying when species goes from absent to present in survey
data_ce_col <- holding %>% 
  filter(Diff == 1) %>% 
  group_by(FishRegionsurvey, spp) %>% 
  summarise(col = n())

# Identifying when species goes from present to absent in survey
data_ce_ext <- holding %>% 
  filter(Diff == -1) %>% 
  group_by(FishRegionsurvey, spp) %>% 
  summarise(ext = n())

# Identifying when species remains as either absent or present in survey
data_ce_stable <- holding %>% 
  filter(Diff == 0) %>% 
  group_by(FishRegionsurvey, spp) %>% 
  summarise(stable = n())

# merging three objects above
data_ce <- data_ce_stable %>%
  full_join(data_ce_col, by = c("FishRegionsurvey", "spp")) %>% 
  full_join(data_ce_ext, by = c("FishRegionsurvey", "spp")) %>% 
  replace(is.na(.), 0) %>% 
  select(-stable) %>% 
  left_join(spp_list, by = c("FishRegionsurvey", "spp")) %>% 
  select(FishRegionsurvey, spp, common, col, ext)
```

```{r build dataframe}
#adding colonization and extinction event columns to data frame to use in trend categorization
data <- data_sum_All %>% 
  left_join(data_ce, by = c("FishRegionsurvey", "spp", "common"))

```

```{r add_trend_category function}

add_trend_category <- function(data){
    data %>%
    arrange(region, FishRegionsurvey, spp, common, year) %>%
    group_by(region, FishRegionsurvey, spp, common) %>%
    group_modify(~{
      df <- .x
      n_years <- nrow(df)
      abs_vec <- as.numeric(df$Absence_Presence)
      
      # 1. Basic Stats
      percent_present <- mean(abs_vec, na.rm = TRUE)
      
      # 2. Windowed Stats (First 1/3 vs Last 1/3)
      third_n <- max(1, floor(n_years / 3))
      first_third <- abs_vec[1:third_n]
      last_third  <- abs_vec[(n_years - third_n + 1):n_years]
      
      mean_first <- mean(first_third, na.rm = TRUE)
      mean_last  <- mean(last_third, na.rm = TRUE)
      
      # 3. Recent Window (Last 5 years)
      last5_idx <- (max(1, n_years - 4)):n_years
      recent_5  <- abs_vec[last5_idx]
      sum_recent_5 <- sum(recent_5, na.rm = TRUE)
      
      # 4. New logic for the last 10 years
      last10_n <- min(10, n_years)
      last10_idx <- (n_years - last10_n + 1):n_years
      recent_10 <- abs_vec[last10_idx]
      mean_recent_10 <- mean(recent_10, na.rm = TRUE)
      
      # 3. Recent Window (Last 10 years)
      last10_idx <- (max(1, n_years - 9)):n_years
      recent_10  <- abs_vec[last10_idx]
      sum_recent_10 <- sum(recent_10, na.rm = TRUE)
      
      # 5. Run Length Encoding (Streaks)
      # Calculates the length of the final 'streak' of 1s or 0s
      runs <- rle(abs_vec)
      last_run_val <- tail(runs$values, 1)
      last_run_len <- tail(runs$lengths, 1)
      
      # 5. Logic Hierarchy (Established vs. Trending)
      trend_category <- case_when(
        
        # 1. PERSISTENTLY PRESENT
        # Species that were ALWAYS high (not a new establishment)
        # and any absences are not a recent streak of 3+ absences
        percent_present >= 0.85 & !(last_run_val == 0 & last_run_len >= 3)
        ~"Persistently Present",
          
        # 2, ABSENT / VAGRANT
        # It was very rarely seen (percent_present <= 0.05)
        # and if caught the hits aren't a recent streak of presences 
        percent_present <= 0.05 & !(last_run_val == 1 & last_run_len >= 3) ~ "Absent",
        
        # 3. ESTABLISHED / SUSTAINED (Emerged + Stabilized)
        # Low start/Sporadic start that is now high and consistent
        #A species that has "stabilized" usually satisfies two conditions:
          #High recent occupancy: They are present in almost all of the most recent years
          #Low early occupancy: They were infrequent or absent in the beginning of the time series.
        (mean_first <= 0.50 & mean_last >= 0.85 & last_run_val == 1 & last_run_len >=5) ~ "Established",
        
        # 4. POTENTIALLY EMERGING
        # Trend is upward, caught recently, but not 'solid' yet
        (mean_first <= 0.50 & mean_first < mean_last & sum_recent_5 >= 4) ~ "Potentially Emerging",
        
        # 5. DISSAPPEARED / SUSTAINED ABSENCE 
        # high start, but gone now
        (mean_first >= 0.50 & mean_last <= 0.15 & last_run_val == 0 & last_run_len >=5)
        ~ "Dissappeared",
        
        # 6. POTENTIALLY DISAPPEARING
        (mean_first >= 0.50 & mean_first > mean_last & sum_recent_5 <= 1) ~ "Potentially Disappearing",
        
        # 7. SPORADIC
        TRUE ~ "Sporadic"
      )
      
      df$TrendCategory <- trend_category
      return(df)
    }) %>%
    ungroup()
}

# This dataframe will be used for the final data exporting
data_summary <- add_trend_category(data) %>%
  select(region, survey, FishRegion, FishRegionsurvey, spp, common, year, FMCwtcpue, Totwtcpue, prop, Absence_Presence, TrendCategory) %>%
  distinct() %>% 
  mutate(FishRegionsurvey = as.character(FishRegionsurvey),
         spp = as.character(spp))

##Trend summaries
category_summary <- data_summary %>%
  group_by(FishRegionsurvey, TrendCategory) %>%
  summarize(species_count = n_distinct(spp), .groups = "drop") %>%
  arrange(FishRegionsurvey, desc(species_count))

print(category_summary)

#visualize, check, validate
#Validation
# 1. Select one representative species for each category
examples <- data_summary %>%
  group_by(TrendCategory) %>%
  slice_sample(n = 1) %>%
  pull(common)

# 2. Filter data for these specific examples
plot_data <- data_summary %>%
  filter(common %in% examples) %>%
  mutate(Absence_Presence = factor (Absence_Presence, levels = c(0,1)))

# 3. Create the Validation Plot
ggplot(plot_data, aes(x = year, y = common, fill = Absence_Presence)) +
  geom_tile(color = "white", linewidth = 0.2) + # Adds a small border between tiles
  facet_wrap(~TrendCategory, scales = "free_y", ncol = 1) + # Group by TrendCategory
  scale_fill_manual(
    values = c("0" = "#f8d7da", "1" = "#004085"), # Light red vs Deep blue
    labels = c("Absent", "Present"),
    name = "Status"
  ) +
  theme_minimal() +
  labs(
    title = "Species Trend Validation: Occupancy Barcodes",
    subtitle = "Visualizing the 'Sustained' vs 'Sporadic' patterns",
    x = "Year",
    y = "Species Common Name"
  ) +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    panel.grid = element_blank(), # Remove grid lines for a cleaner tile look
    axis.text.y = element_text(size = 8)
  )
```

```{r}
## Looking Across surveys for individual species view
# 1. Select the species you want to investigate
# You can change these names to the specific fish you are interested in
selected_spp <- c("Centropristis striata") 

plot_data <- data_summary %>%
  filter(spp %in% selected_spp) %>%
  # Ensure the status is a factor for the discrete color scale
  mutate(Absence_Presence = factor(Absence_Presence, levels = c(0, 1)))

# 2. Create the Plot
ggplot(plot_data, aes(x = year, y = paste(FishRegionsurvey, "-", TrendCategory), fill = Absence_Presence)) +
  geom_tile(color = "white", linewidth = 0.1) +
  # Facet by species name, allowing each species to have its own block
  facet_wrap(~common, ncol = 1, scales = "free_y") + 
  scale_fill_manual(
    values = c("0" = "#f8d7da", "1" = "#004085"), # Light red (absent) vs Deep blue (present)
    labels = c("Absent", "Present"),
    name = "Status"
  ) +
  theme_minimal() +
  labs(
    title = "Regional Occupancy Barcodes by Species",
    subtitle = "Comparing presence/absence patterns across different fish surveys",
    x = "Year",
    y = "Fish Region Survey"
  ) +
  theme(
    # Make the species headers bold and distinct
    strip.text = element_text(face = "bold", size = 12, color = "darkslategrey"),
    # Clean up the background
    panel.grid = element_blank(),
    # Adjust survey labels to be readable
    axis.text.y = element_text(size = 8),
    legend.position = "bottom",
    # Add spacing between the species facets
    panel.spacing = unit(1, "lines")
  )

```

## Increasing vs. decreasing vs. stable trends in proportion biomass
This section is caculates trend in proportion within each FMC and adds to dataframe

```{r packages}
library(trend)
```

```{r prop inc vs dec}
# 1. Define a function to categorize the trend
get_trend_status <- function(prop_vector) {
  # 1. Remove NAs for the calculation
  clean_vec <- na.omit(prop_vector)
  
  # 2. Check for minimum data points (need at least 5 for MK test)
  if(length(clean_vec) < 5) return("Incomplete Data")
  
  # 3. Check if the values are constant (e.g., all 0 or all 1)
  # If the difference between max and min is ~0, it's stable/zero
  if(max(clean_vec) - min(clean_vec) < 1e-9) return("Stable (Zero/Constant)")
  
  # 4. Run the test now that we know there is variation
  test <- trend::mk.test(clean_vec)
  p <- test$p.value
  s <- test$estimates[1] 
  
  if (p < 0.05) {
    if (s > 0) return("Significant Increase")
    else return("Significant Decrease")
  } else {
    return("Stable")
  }
}

# 2. Apply to the dataframe and create a new column
data_summary <- data_summary %>%
  group_by(spp, FishRegionsurvey) %>%
  mutate(trend_status = get_trend_status(prop)) %>%
  ungroup()

#add a shorthand trend name
data_summary <- data_summary %>%
  mutate(Trend = case_when(
    trend_status == "Significant Increase" ~ "↑ Inc",
    trend_status == "Significant Decrease" ~ "↓ Dec",
    TRUE ~ "Stable"
  ))


```

```{r}
#Plot the data with the trends
##selected subset of species
selected_spp <- c("Centropristis striata") 
plot_data_spring <- data_summary %>%
  filter(spp %in% selected_spp) %>%
  # Ensure the status is a factor for the discrete color scale
  mutate(FishRegion = factor(FishRegion, 
                               levels = c("New England", "NEMA", "South Atlantic"))) %>%
  filter(survey=="NEFSC Spring Bottom Trawl")

# Convert year to numeric
plot_data_spring$year <- as.numeric(as.character(plot_data_spring$year)) 

## create a summary table for the trends info 
library(ggpubr) # Great for arranging tables and plots

# 1. Create the summary table data
summary_tab <- plot_data_spring %>%
  distinct(FishRegion, TrendCategory, Trend) 

# 2. Turn that data into a clean-looking table object
table_plot <- ggtexttable(summary_tab, rows = NULL, theme = ttheme("default"))

##plot the data
color_palette <- scale_fill_manual(
  name = "Jurisdiction", # Legend title
  values = c("#275F6E","#3B8EA5","#D1E3F0")) # Example colors: Blue, Dark Blue, Off-White


barplot <- ggplot(plot_data_spring, aes(x = year, y = prop, fill = FishRegion)) +
  geom_bar(stat = "identity", position = "stack", width = 1) +
   # Color palette and legend
  color_palette +
    # Add the trend labels
  # geom_text(data = trend_labels, 
  #           aes(x = mid_year, y = 0.5, label = label_text), # y=0.5 places it in the middle height
  #           color = "white", 
  #           fontface = "bold",
  #           size = 3,
  #           inherit.aes = FALSE) + # Important: prevents geom_text from looking for 'fill'
  facet_wrap(~ spp, ncol = 2) +
  labs(title = "Biomass Proportions with Significant Trends (p < 0.05)",
       x = "Year", y = "Proportion") +
  theme(
    strip.text = element_text(hjust = 0), # Align species titles left
    legend.position = "bottom",
    axis.text.x = element_text(angle = 0)
  )

final_output <- ggarrange(barplot, table_plot, ncol=1, heights=c(2,1))

```
##Data Expore: NEFSC FMC trends 
```{r NEFSC data export}
write.csv(data_summary, here::here("data_processing_rcode","output","data_clean","NEFSC_FMC_SpeciesPersistenceIndicatorTrend.csv"))
```

